---
title: "Introducing `adanear`"
output:
  rmarkdown::html_vignette:
  toc: true
bibliography: ../inst/references/citation.bib
link-citations: true
nocite: |
  @brito2024framework
  @he2008adasyn
  @mani2003knn
  @hvitfeldt2025themis
  @kuhn2025tidymodels
vignette: >
  %\VignetteIndexEntry{Introducing adanear}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

`adanear` provides efficient binary class balancing steps for `recipes` and the broader `tidymodels` ecosystem.

The first implemented method is `step_adanear()`, a hybrid approach that combines ADASYN style oversampling with NearMiss style undersampling.

The package was designed for binary classification workflows in which the minority class is rare and model performance depends strongly on how the training data are rebalanced before fitting.

# Motivation

The package is motivated by the banking churn framework proposed by @brito2024framework.

ADASYN allocates synthetic samples according to local learning difficulty @he2008adasyn.

NearMiss is an informed undersampling strategy based on nearest neighbors @mani2003knn.

For comparison, `themis` provides `step_adasyn()` and `step_nearmiss()` as separate recipe steps @hvitfeldt2025themis.


# The `step_adanear()` approach

`step_adanear()` combines two ideas.

First, it uses ADASYN style oversampling to generate synthetic minority observations in more difficult regions of the feature space.

Second, it uses NearMiss style undersampling to retain majority observations closer to the minority boundary.

This hybrid structure aims to improve training efficiency and sharpen the decision boundary without relying on purely random oversampling or purely random undersampling.

At a high level, the step:

1. validates the binary outcome and numeric predictors  
2. normalizes predictors  
3. estimates local difficulty for minority observations  
4. generates synthetic minority rows by interpolation  
5. restores binary and integer-like predictor types  
6. applies NearMiss style selection to the majority class  
7. returns a balanced training set for downstream modeling.

# Mathematical intuition

## ADASYN

ADASYN allocates more synthetic examples to minority observations that are harder to learn.

Let:

-   $m_s$ be the number of minority instances
-   $m_l$ be the number of majority instances
-   $\beta \in [0, 1]$ control the target amount of new synthetic data
-   $G$ be the total number of synthetic minority instances to generate.

A common formulation is:

$$
  G = (m_l - m_s)\beta
$$

For each minority point $x_i$, define a local difficulty score $r_i$ based on the proportion of majority neighbors among its $k$ nearest neighbors:

$$
  r_i = \frac{\Delta_i}{k}
$$

where $\Delta_i$ is the number of majority neighbors among the $k$ nearest neighbors of $x_i$.

These scores are normalized into weights:

$$
  \hat{r}_i = \frac{r_i}{\sum_{j=1}^{m_s} r_j}
$$

The number of synthetic points allocated to minority point $x_i$ is then:

$$
  g_i = \hat{r}_i G
$$

A new synthetic point is generated by interpolation:

$$
  x_{new} = x_i + \lambda (x_{zi} - x_i), \quad \lambda \sim U(0,1)
$$

where $x_{zi}$ is a randomly selected minority neighbor of $x_i$.

The intuition is that observations in more ambiguous local neighborhoods receive more synthetic support.

## NearMiss

NearMiss is an informed undersampling strategy for the majority class.

Let $x_j^{maj}$ denote a majority point. For each majority point, compute the average distance to its $k$ nearest minority neighbors:

$$
  d_j = \frac{1}{k}\sum_{t=1}^{k} \delta(x_j^{maj}, x_t^{min})
$$

where $\delta(\cdot,\cdot)$ is a distance function, typically Euclidean distance.

The retained majority points are those with smaller average distance to the minority class, since these are more informative near the decision boundary.

This makes NearMiss fundamentally different from random undersampling, which may remove relevant majority observations and weaken the learned boundary.

# Why combine ADASYN and NearMiss

ADASYN and NearMiss act on different sides of the class imbalance problem.

ADASYN increases representation in difficult minority regions, while NearMiss reduces majority redundancy in regions closest to the minority class.

Combining them can be useful because ADASYN reinforces underrepresented minority neighborhoods, NearMiss concentrates the majority class near the regions that matter most, the final training distribution can become more informative for classification, and the downstream learner can receive a better representation of the event boundary.

# Positioning within tidymodels

`adanear` was designed to integrate naturally with `recipes`, `workflows`, `parsnip`, `rsample`, and `yardstick`.

The step follows the standard architecture of a custom `recipes` step:

-   `step_adanear()`
-   `prep.step_adanear()`
-   `bake.step_adanear()`
-   `print.step_adanear()`
-   `tidy.step_adanear()`

This means it can be inserted in the same place where users would already place other preprocessing operations in a `recipe`.

# Basic example with tidymodels

The example below uses `modeldata::two_class_dat()` to illustrate a full binary classification workflow with `step_adanear()` and `xgboost` under the `binary:logistic` objective.

```{r basic-example}
library(tidymodels)
library(adanear)

set.seed(123)

dados <- modeldata::two_class_dat(3000) |>
  dplyr::mutate(Class = factor(Class))

splitObj <- rsample::initial_split(dados, strata = Class)
dadosTreino <- rsample::training(splitObj)
dadosTeste <- rsample::testing(splitObj)

rec <- recipes::recipe(Class ~ ., data = dadosTreino) |>
  step_adanear(
    Class,
    increaseRatio = 0.20,
    neighborsAdasyn = 5L,
    neighborsNearMiss = 5L,
    nThreads = 1L,
    seed = 42L
  )

modelo <- parsnip::boost_tree(
  mode = "classification",
  trees = 300,
  learn_rate = 0.05,
  tree_depth = 6,
  min_n = 5
) |>
  parsnip::set_engine(
    engine = "xgboost",
    objective = "binary:logistic",
    eval_metric = "aucpr"
  )

wf <- workflows::workflow() |>
  workflows::add_recipe(rec) |>
  workflows::add_model(modelo)

folds <- rsample::vfold_cv(dadosTreino, v = 5, strata = Class)

resultado <- tune::fit_resamples(
  wf,
  resamples = folds,
  metrics = yardstick::metric_set(pr_auc, roc_auc, accuracy)
)

tune::collect_metrics(resultado)
```

# Inspecting the step

Like other `recipes` steps, `step_adanear()` supports standard inspection via `prep()`, `tidy()`, and `print()`.

```{r inspect-step}
recPreparada <- recipes::prep(rec, training = dadosTreino)

recPreparada
recipes::tidy(recPreparada, number = 1)
```

# Where to place `step_adanear()` in a recipe

In practice, `step_adanear()` should be placed after operations that make the predictor space suitable for distance-based computations.

Typical upstream preprocessing may include:

- missing value imputation  
- unknown level handling  
- rare category collapsing  
- dummy encoding  
- removal of zero variance predictors.

A common pattern is:

```{r recipe-pattern, eval = FALSE}
recipes::recipe(target ~ ., data = dadosTreino) |>
  recipes::step_impute_median(recipes::all_numeric_predictors()) |>
  recipes::step_unknown(recipes::all_nominal_predictors()) |>
  recipes::step_other(recipes::all_nominal_predictors(), threshold = 0.01) |>
  recipes::step_dummy(recipes::all_nominal_predictors(), one_hot = TRUE) |>
  recipes::step_nzv(recipes::all_predictors()) |>
  step_adanear(target)
```

Since the step operates on distances, its intended usage assumes that the predictors used in the balancing stage are numeric, finite, and free of missing values.

# Comparison with themis

The `themis` package offers separate balancing steps such as `step_adasyn()` and `step_nearmiss()`.

This makes `themis` a natural baseline when evaluating `adanear` in a `recipes` based workflow.

A fair comparison should be made by holding constant:

- the training split  
- the resampling design  
- the downstream model specification  
- the evaluation metrics  
- the computational environment.

The code below provides a reproducible benchmark template.

## Timing benchmark

```{r benchmark-timing, eval = FALSE}
library(tidymodels)
library(themis)
library(adanear)
library(bench)

set.seed(42L)

dados <- modeldata::two_class_dat(5000) |>
  dplyr::mutate(Class = factor(Class))

splitObj <- rsample::initial_split(dados, strata = Class)
dadosTreino <- rsample::training(splitObj)

recSattva <- recipes::recipe(Class ~ ., data = dadosTreino) |>
  step_adanear(
    Class,
    increaseRatio = 0.20,
    neighborsAdasyn = 5L,
    neighborsNearMiss = 5L,
    nThreads = 1L,
    seed = 123L
  )

recThemis <- recipes::recipe(Class ~ ., data = dadosTreino) |>
  themis::step_adasyn(Class, over_ratio = 1) |>
  themis::step_nearmiss(Class, under_ratio = 1)

benchResultado <- bench::mark(
  adanear = {
    recipes::prep(recSattva, training = dadosTreino) |>
      recipes::bake(new_data = dadosTreino)
  },
  themis = {
    recipes::prep(recThemis, training = dadosTreino) |>
      recipes::bake(new_data = dadosTreino)
  },
  iterations = 10,
  check = FALSE
)

benchResultado
```

## Downstream performance benchmark

```{r benchmark-performance, eval = FALSE}
modelo <- parsnip::boost_tree(
  mode = "classification",
  trees = 300,
  learn_rate = 0.05,
  tree_depth = 6,
  min_n = 5
) |>
  parsnip::set_engine(
    "xgboost",
    objective = "binary:logistic",
    eval_metric = "aucpr"
  )

wfSattva <- workflows::workflow() |>
  workflows::add_recipe(recSattva) |>
  workflows::add_model(modelo)

wfThemis <- workflows::workflow() |>
  workflows::add_recipe(recThemis) |>
  workflows::add_model(modelo)

folds <- rsample::vfold_cv(dadosTreino, v = 5, strata = Class)

resSattva <- tune::fit_resamples(
  wfSattva,
  resamples = folds,
  metrics = yardstick::metric_set(pr_auc, roc_auc, accuracy)
)

resThemis <- tune::fit_resamples(
  wfThemis,
  resamples = folds,
  metrics = yardstick::metric_set(pr_auc, roc_auc, accuracy)
)

tune::collect_metrics(resSattva)
tune::collect_metrics(resThemis)

```

# Interpreting the comparison

The comparison with `themis` should be interpreted carefully.

If `adanear` is faster on your machine, that should appear in the benchmark statistics returned by `bench::mark()`, especially median runtime and memory allocation.

If `adanear` also yields better downstream predictive performance, that should appear in metrics such as PR AUC, ROC AUC, or recall under the same resampling design and model specification.

It is good practice not to claim superiority before reproducing these benchmarks in the target environment.

# Recommended evaluation metrics

Under severe imbalance, some metrics are more informative than others.

## PR AUC

PR AUC is often especially useful when the positive class is rare because it emphasizes precision and recall.

## ROC AUC

ROC AUC remains useful as a ranking metric, but it can look optimistic when the class imbalance is strong.

## Accuracy

Accuracy should be interpreted with caution, since a model can obtain high accuracy by predicting mostly the majority class.

## Recall and specificity

Recall for the event class and specificity for the non-event class are useful complementary summaries.

# Practical notes

## Reproducibility

Since `step_adanear()` contains stochastic elements, fixing a `seed` is recommended whenever reproducibility matters.

## Parallelism

The current implementation exposes `nThreads` to control internal computation. For reproducible benchmarking, it is often preferable to set `nThreads = 1L` first.

## Type restoration

The step includes internal logic to restore binary and integer-like predictors after synthetic generation, which is important for practical workflows with discrete inputs.

## Skip behavior

As with other supervised sampling steps, the intended use is training-time balancing inside the recipe. The step should be configured so that it affects the training data and not future assessment data inappropriately.

# Minimal reproducible workflow

For quick experimentation, the shortest useful workflow is:

```{r minimal-workflow, eval = FALSE}
library(tidymodels)
library(adanear)

dados <- modeldata::two_class_dat(2000)

wf <- workflows::workflow() |>
  workflows::add_recipe(
    recipes::recipe(Class ~ ., data = dados) |>
      step_adanear(Class)
  ) |>
  workflows::add_model(
    parsnip::boost_tree(mode = "classification") |>
      parsnip::set_engine("xgboost", objective = "binary:logistic")
  )

folds <- rsample::vfold_cv(dados, v = 5, strata = Class)

tune::fit_resamples(
  wf,
  resamples = folds,
  metrics = yardstick::metric_set(pr_auc, roc_auc)
)
```

# Limitations and scope of the first version

This first version of `adanear` focuses on a single binary balancing step.

At this stage, the package is intentionally scoped to:

-   binary classification,
-   `recipes` based preprocessing,
-   numeric predictor spaces suitable for distance-based balancing,
-   reproducible modeling workflows in `tidymodels`.

Future versions may add:

-   standalone ADASYN and NearMiss steps
-   additional hybrid balancing methods
-   multiclass variants
-   richer audit outputs
-   expanded benchmarking material.

# Session info

```{r session-info}
sessionInfo()
```

# References
